---
title: "Data Carpentry in R - exercises"
output: 
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: pygments
    df_print: kable
    code_folding: hide
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

[back to course materials](https://datacarpentry.org/R-ecology-lesson/index.html)

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
surveys <- read_csv("https://ndownloader.figshare.com/files/2292169")
```


# Intro

----

## Variables

What is the value of `mass_index` after running these commands?

```
mass <- 47.5
age  <- 122
mass_index <- mass/age
mass <- mass * 2.0
```

What type of vector do you get in each of the following cases (hint: use `class()`)?

```
num_char <- c(1, 2, 3, "a")
num_logical <- c(1, 2, 3, TRUE)
char_logical <- c("a", "b", "c", TRUE)
tricky <- c(1, 2, 3, "4")
```

----

## Conditional subsetting

Using this vector:

```
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

```{r, echo=FALSE}
heights <- c(63, 69, 60, 65, NA, 68, 61, 70, 61, 59, 64, 69, 63, 63, NA, 72, 65, 64, 70, 63, 65)
```

* create a new vector with missing values removed

```{r, results=FALSE}
heights_no_na <- heights[!is.na(heights)]
```

* calculate the median of heights

```{r, results=FALSE}
median(heights, na.rm = TRUE)
# or
median(heights_no_na)
```

* calculate how many values are above 67

```{r, results=FALSE}
sum(heights_no_na > 67)
```

----

## Basic `data.frame` manipulation

* How many rows and columns does `surveys` have? (hint: functions `ncol()` and `nrow()`)

```{r, results=FALSE}
nrow(surveys)
ncol(surveys)
```

* What type of variables do you have in the table? (hint: function `glimpse()`)

```{r, results=FALSE}
glimpse(surveys)
```

4. How many missing values of weight are there? (remember you can use `sum()` on a logical vector)
    * extra: what proportion of values are missing?
    
```{r, results=FALSE}
sum(is.na(surveys$weight))

# Extra: divide by total number of observations
sum(is.na(surveys$weight))/nrow(surveys)
```

5. How many unique genera were sampled? (hint: functions `unique()` and `length()`)

```{r, results=FALSE}
length(unique(surveys$genus))
```


# Manipulating data frames with `dplyr`

----

## filter and select

Using pipes, subset the surveys data to include animals collected before 1995 and 
retain only the columns year, sex, and weight. 

```{r}
surveys_pre1995 <- surveys %>% 
  filter(year < 1995) %>% 
  select(year, sex, weight)
```

The answer should have `r nrow(surveys_pre1995)` rows of data.

----

## mutate, filter and select

Create a new data frame from the surveys data that meets the following criteria: 
contains only the `species_id` column and a new column called `hindfoot_half` 
containing values that are half the `hindfoot_length` values. 
In this `hindfoot_half` column, there are no `NAs` and all values are less than 30.

Hint: think about how the commands should be ordered to produce this data frame!

```{r}
surveys_hindfoot_half <- surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    mutate(hindfoot_half = hindfoot_length / 2) %>%
    filter(hindfoot_half < 30) %>%
    select(species_id, hindfoot_half)
```

The answer should have `r nrow(surveys_hindfoot_half)` rows and `r ncol(surveys_hindfoot_half)` columns.


----

## grouped summaries

* How many animals were caught in each `plot_type` surveyed?

```{r, results=FALSE}
surveys %>%
    count(plot_type) 
```

* Make a summary table with the mean, minimum and maximum hindfoot length of each.
Include the total number of individuals observed as well as the number of those 
individuals that had data for this variable.

```{r, results=FALSE}
surveys %>%
    filter(!is.na(hindfoot_length)) %>%
    group_by(species_id) %>%
    summarize(
        mean_hindfoot_length = mean(hindfoot_length),
        min_hindfoot_length = min(hindfoot_length),
        max_hindfoot_length = max(hindfoot_length),
        n = n(),
        n_hindfoot = sum(!is.na(hindfoot_length))
    )
```

* What was the heaviest animal measured in each year? Return the columns `year`, 
`genus`, `species_id` and `weight`.

```{r, results=FALSE}
surveys %>%
    filter(!is.na(weight)) %>%
    group_by(year) %>%
    filter(weight == max(weight)) %>%
    select(year, genus, species, weight) %>%
    arrange(year)
```


----

## Data cleaning

* Create a new data frame called `surveys_complete`, which:
    * has no missing values for `weight`, `hindfoot_length` and `sex`
    * contains only species that were measured in at least 50 individuals

```{r}
surveys_complete <- surveys %>% 
  filter(!is.na(weight) & !is.na(sex) & !is.na(hindfoot_length))

common_species <- surveys_complete %>% 
  count(species_id) %>% 
  filter(n >= 50)

surveys_complete <- surveys_complete %>% 
  filter(species_id %in% common_species$species_id)
```

The final data frame should have `r nrow(surveys_complete)` rows.

* Optional: save the new table as a CSV file in the `data_output` folder of your 
project directory.

```{r, eval=FALSE}
write_csv(surveys_complete, "data_output/surveys_complete.csv")
```


# Plotting with `ggplot2`

----

## distributions

Boxplots are useful summaries, but hide the shape of the distribution. 
For example, if the distribution is bimodal, we would not see it in a boxplot. 
An alternative to the boxplot is the violin plot, where the shape 
(of the density of points) is drawn.

* Use `geom_violin()` to create violin plots of weight distributions for each `genus`

```{r, results=FALSE}
surveys_complete %>% 
  ggplot(aes(genus, weight)) +
  geom_violin()
```

* Some individuals are much heavier than others. Using a log scale in this case would help.
Try adding a "scale" layer with `scale_y_log10()`. 

```{r, results=FALSE}
surveys_complete %>% 
  ggplot(aes(genus, weight)) +
  geom_violin() +
  scale_y_log10()
```

* Can you overlay a boxplot with outliers removed (check `?geom_boxplot` help)?

```{r, results=FALSE}
surveys_complete %>% 
  ggplot(aes(genus, weight)) +
  geom_violin() +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  scale_y_log10()
```

* Try and recreate the picture below, showing the distribution of weights by `species_id`,
with violin plots coloured by `genus`.

```{r}
surveys_complete %>% 
  ggplot(aes(species_id, weight)) +
  geom_violin(aes(fill = genus)) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  scale_y_log10()
```

----

## Two categorical variables

Use the `count()` function to get the number of individuals of each genus in 
each plot type surveyed and then use that table to make a visual representation 
like the one shown below. 

(hint: try the `size` aesthetic with `geom_point()`)

```{r}
surveys_complete %>%
  count(genus, plot_type) %>% 
  ggplot(aes(genus, plot_type)) +
  geom_point(aes(size = n)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Genus", y = "Plot Type")
```

----

## Line plots

We want to explore if the species diversity (number of different species) changed 
across the years in different plot treatments (`plot_type`). 

* Make a new table called `yearly_diversity`, with the number of _distinct_ 
`species_id` sampled in each year and plot type. 

(hint: use the function `n_distinct()` inside the `summarise()` function)

```{r}
yearly_diversity <- surveys_complete %>% 
  group_by(year, plot_type) %>% 
  summarise(n_species = n_distinct(species_id)) %>% 
  ungroup()
```
  
* Using the new table, try and reproduce the plot below

```{r}
yearly_diversity %>% 
  ggplot(aes(year, n_species)) +
  geom_line(aes(colour = plot_type)) +
  labs(colour = "Plot Type", x = "Year", y = "# species")
```

* Update your summary table by also including information about sex (i.e. species 
diversity per `year`, `plot_type` and `sex`). Can you think of ways to change 
your previous plot to also visualise this extra information?

```{r, echo=FALSE}
# Add `sex` to the group_by call
yearly_diversity <- surveys_complete %>% 
  group_by(year, plot_type, sex) %>% 
  summarise(n_species = n_distinct(species_id)) %>% 
  ungroup()

# Improve visualisation by faceting the plot horizontally
yearly_diversity %>% 
  ggplot(aes(year, n_species)) +
  geom_line(aes(colour = plot_type)) +
  labs(colour = "Plot Type", x = "Year", y = "# species") +
  facet_grid(sex ~ .)
```


----

## Factors to improve visualisation

* Using your `yearly_diversity` table, can you recreate the plot below, ensuring 
the order of the `plot_type` facets is as shown?

```{r}
# Make a vector with treatments in desired order
## this is optional, but makes it easier to read the code
plot_type_order <- c("Control", 
                     "Short-term Krat Exclosure", "Long-term Krat Exclosure",
                     "Rodent Exclosure", "Spectab exclosure")

# Factorise the `plot_type` variable with custom levels before plotting
yearly_diversity %>% 
  mutate(plot_type = factor(plot_type, levels = plot_type_order)) %>% 
  ggplot(aes(year, n_species)) +
  geom_line(aes(colour = sex)) +
  labs(colour = "Plot Type", x = "Year", y = "# species") +
  facet_wrap( ~ plot_type)
```


# Optional (advanced) plotting exercises

----

## Line plots

Use line plots to help you explore these questions:

* How did the average weight of all species change across the years?

```{r, results=FALSE}
surveys_complete %>% 
  group_by(year) %>% 
  summarise(mean_wgt = mean(weight, na.rm = TRUE)) %>% 
  ggplot(aes(year, mean_wgt)) +
  geom_line() + geom_point() +
  labs(x = "Year", y = "Average biomass (g)")
```

* Was it different for males and females? 

```{r, results=FALSE}
surveys_complete %>% 
  group_by(year, sex) %>% 
  summarise(mean_wgt = mean(weight, na.rm = TRUE)) %>% 
  ggplot(aes(year, mean_wgt, colour = sex)) +
  geom_line() + geom_point() + 
  labs(x = "Year", y = "Average biomass (g)")
```

* What about between different `plot_type`? (hint: use facetting to help you)

```{r, results=FALSE}
surveys_complete %>% 
  group_by(year, sex, plot_type) %>% 
  summarise(mean_wgt = mean(weight, na.rm = TRUE)) %>% 
  ggplot(aes(year, mean_wgt, colour = sex)) +
  geom_line() + geom_point() + 
  facet_wrap(~ plot_type) +
  labs(x = "Year", y = "Average biomass (g)")
```

* Bonus: Try and add a "ribbon" which shows the mean +/- 2*SEM (standard error of 
the mean, calculated as the standard deviation divided by the square root of the 
number of observations)

```{r}
surveys_complete %>% 
  group_by(year, sex, plot_type) %>% 
  summarise(mean_wgt = mean(weight),
            se_wgt = sd(weight)/sqrt(n())) %>% 
  ggplot(aes(year, mean_wgt, colour = sex)) +
  geom_ribbon(aes(ymin = mean_wgt - 2*se_wgt, ymax = mean_wgt + 2*se_wgt, fill = sex), 
              alpha = 0.3, colour = NA) +
  geom_line() + 
  facet_wrap( ~ plot_type) +
  labs(x = "Year", y = "Average biomass (g)")
```


----

## Reshaping data

* Make a scatterplot to explore the correlation between average male and female 
weights per species in each `plot_type` surveyed. (hint: add a x=y identity line 
using `geom_abline()`)

```{r, warning=FALSE}
surveys_complete %>% 
  group_by(species_id, plot_type, sex) %>% 
  summarise(mean_wgt = mean(weight)) %>% 
  ungroup() %>% 
  spread(sex, mean_wgt) %>% 
  ggplot(aes(M, F)) +
  geom_abline(linetype = 2) +
  geom_point(aes(colour = factor(plot_type))) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Male", y = "Female", colour = "Plot type")
```


----

## Reordering factors

* Make a boxplot similar to the one below, showing the hindfoot length distribution, 
with the x-axis sorted by mean weight. 
(hint: create a colum with the mean weight per species and then use the function 
`reorder()` to factorise the x-axis variable)

```{r}
surveys_complete %>% 
  group_by(species_id) %>% 
  mutate(mean_weight = mean(weight)) %>% 
  ungroup() %>% 
  mutate(species_id = reorder(species_id, mean_weight)) %>% 
  ggplot(aes(species_id, hindfoot_length)) +
  geom_boxplot(aes(fill = log10(mean_weight))) +
  scale_fill_viridis_c() # "viridis" is a colour-blind pallete
```

----

## Combining datasets

In a publication from [Ernest et al. 2017](https://www.journals.uchicago.edu/doi/10.1086/592402) 
some species where classified according to their habitat preferences. 

Read this information into your R session using the following command:

```
habitat_pref <- read_csv("https://raw.githubusercontent.com/tavareshugo/data_carpentry_extras/master/slides_with_exercises/species_habitat_affinities_Ernest2017.csv")
```

```{r, message=FALSE, echo=FALSE}
habitat_pref <- read_csv("https://raw.githubusercontent.com/tavareshugo/data_carpentry_extras/master/slides_with_exercises/species_habitat_affinities_Ernest2017.csv")
```

* How many of these species are contained in your `surveys_complete` dataset? (hint: 
the function `paste()` might be useful)

```{r, results=FALSE}
sum(habitat_pref$species %in% paste(surveys_complete$genus, surveys_complete$species))
```

* Make a boxplot like the one below, showing the weight distribution of each 
species that has habitat affinity information. (hint: look at the help for 
`?facet_grid` to see how to make the panels scales and spacing adjust to the 
data)

```{r}
surveys_complete %>% 
  mutate(species_name = paste(genus, species)) %>% 
  inner_join(habitat_pref, by = c("species_name" = "species")) %>% 
  ggplot(aes(species_id, weight)) +
  geom_boxplot(aes(fill = genus)) +
  facet_grid(sex ~ habitat, scale = "free_x", space = "free_x") +
  scale_y_log10()
```


* Plot the change in abundance across the years for species with different habitat 
affinities (include those species with no habitat information as well). 
    * Abundance = number of individuals per hectare
    * Each plot is 0.25ha

(don't worry too much about getting the exact look of the graph below. But if you
do want to get a similar look, try and do some web search! E.g. "how to change 
legend position in ggplot2")

```{r}
surveys_complete %>% 
  mutate(species_name = paste(genus, species)) %>% 
  left_join(habitat_pref, by = c("species_name" = "species")) %>% 
  group_by(year, habitat) %>% 
  summarise(abundance = n()/(0.25*n_distinct(plot_id))) %>% 
  ggplot(aes(year, abundance)) +
  geom_line(aes(colour = habitat)) +
  labs(x = "Year", y = "Abundance (# individuals/ha)", colour = "Habitat preference") +
  scale_colour_viridis_d(na.value = "grey") +
  theme_classic() +
  theme(legend.position = c(0, 1), legend.justification = c(0, 1))
```


