---
title: "Data Carpentry in R - exercises"
output: 
  slidy_presentation:
    incremental: false
    highlight: monochrome
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
surveys <- read_csv("https://ndownloader.figshare.com/files/2292169")
```


----

* How many animals were caught in each `plot_type` surveyed?
* Make a summary table with the mean, minimum and maximum hindfoot length of each.
Include the total number of individuals observed as well as the number of those 
individuals that had data for this variable.
* What was the heaviest animal measured in each year? Return the columns `year`, 
`genus`, `species_id` and `weight`.


----

* Create a new data frame called `surveys_complete`, which:
    * has no missing values for `weight`, `hindfoot_length` and `sex`
    * contains only species that were measured in at least 50 individuals

```{r, echo=FALSE}
surveys_complete <- surveys %>% 
  filter(!is.na(weight) & !is.na(sex) & !is.na(hindfoot_length))

common_species <- surveys_complete %>% 
  count(species_id) %>% 
  filter(n >= 50)

surveys_complete <- surveys_complete %>% 
  filter(species_id %in% common_species$species_id)
```

The final data frame should have `r nrow(surveys_complete)` rows.


----

Boxplots are useful summaries, but hide the shape of the distribution. 
For example, if the distribution is bimodal, we would not see it in a boxplot. 
An alternative to the boxplot is the violin plot, where the shape 
(of the density of points) is drawn.

* Use `geom_violin()` to create violin plots of weight distributions for each `genus`
* Some individuals are much heavier than others. Using a log scale in this case would help.
Try adding a layer with `scale_y_log10()`. 
* Can you overlay a boxplot with outliers removed (check `?geom_boxplot` help)?
* Try and recreate the picture below, showing the distribution of weights by `species_id`,
with violin plots coloured by `genus`.

```{r, echo = FALSE}
surveys_complete %>% 
  ggplot(aes(species_id, weight)) +
  geom_violin(aes(fill = genus)) +
  geom_boxplot(width = 0.2, outlier.shape = NA) +
  scale_y_log10()
```

----

Make a visual representation of the number of observations we have of each genus 
in each plot type. (hint: try the `size` aesthetic with `geom_point()`)

```{r, echo = FALSE}
surveys_complete %>% 
  ggplot(aes(genus, plot_type)) +
  geom_count() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Genus", y = "Plot Type")
```


----

We want to explore if the species diversity (number of different species) changed 
across the years in different plot treatments (`plot_type`). 

* Make a new table called `yearly_diversity`, with the number of _distinct_ 
`species_id` sampled in each year and plot type. 
(hint: see the function `n_distinct()` to use inside the `summarise()` function)
* Using the new table, try and reproduce the plot below
* Update your summary table by also including information about sex (i.e. species 
diversity per `year`, `plot_type` and `sex`). Can you think of ways to change 
your previous plot to also visualise this extra information?

```{r, echo = FALSE}
surveys_complete %>% 
  group_by(year, plot_type) %>% 
  summarise(n_species = n_distinct(species_id)) %>% 
  ggplot(aes(year, n_species)) +
  geom_line(aes(colour = plot_type)) +
  labs(colour = "Plot Type", x = "Year", y = "# species")
```

----

* Using your `yearly_diversity` table, can you recreate the plot below, but 
change the order of these panels so that M appears on top?

```{r, echo = FALSE}
yearly_diversity <- surveys_complete %>% 
  group_by(year, sex, plot_type) %>% 
  summarise(n_species = n_distinct(species_id))

yearly_diversity %>% 
  ggplot(aes(year, n_species)) +
  geom_line(aes(colour = plot_type)) +
  labs(colour = "Plot Type", x = "Year", y = "# species") +
  facet_grid(sex ~ .)
```


----

Use line plots to help you explore these questions:

* How did the average weight of all species change across the years?
* Was it different for males and females? 
* What about between different `plot_type`? (hint: use facetting to help you)

```{r, eval=FALSE, echo=FALSE}
surveys_complete %>% 
  group_by(year, sex, plot_type) %>% 
  summarise(mean_wgt = mean(weight, na.rm = TRUE)) %>% 
  ggplot(aes(year, mean_wgt, colour = sex)) +
  geom_line() + geom_point() + facet_wrap(~ plot_type)
```


----

* Make a scatterplot to explore the correlation between average male and female 
weights per species in each `plot_type` surveyed. (hint: add a x=y identity line 
using `geom_abline()`)

```{r, echo=FALSE, warning=FALSE}
surveys %>% 
  filter(!is.na(weight) & !is.na(sex)) %>% 
  group_by(species_id, plot_type, sex) %>% 
  summarise(mean_wgt = mean(weight)) %>% 
  ungroup() %>% 
  spread(sex, mean_wgt) %>% 
  ggplot(aes(M, F)) +
  geom_abline(linetype = 2) +
  geom_point(aes(colour = factor(plot_type))) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Male", y = "Female", colour = "Plot type")
```



