---
title: "Syntax equivalents: base R vs Tidyverse"
author: "Hugo Tavares"
date: "20 September 2018"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, results = FALSE)
```


This document is aimed at beggining R users that start by learning "tidyverse" functions. 
It shows how some of the tasks done with "tidyverse" functions have a corresponding 
solution using "base R" syntax (using functions that are part of the core packages 
deployed with R). 

It is important to understand that there are many different ways to achieve the 
same thing, particularly in base R. I have tried to use simpler syntax where possible. 
For the more complicated cases, I have based my solutions on top hits from web 
searches. 

It is also worth remembering that when you look for help online, you might find 
solutions using "base R" syntax. If you are used to "tidyverse" packages, it 
will help to add keywords such as "dplyr" and "tidyverse" to your search.

There are many other resources available online, in particular several cheatsheets 
that might be useful as a reference: https://www.rstudio.com/resources/cheatsheets/ 


# Setup

Load the packages needed for these examples:

```{r, message=FALSE}
library(dplyr)
library(tidyr)
```

The datasets used in these examples come pre-loaded with R. To find more about them 
check their help pages:

```{r}
?iris
?mtcars
?Indometh
```


# Extract variables (columns)

**tidyverse**

```{r}
select(iris, Species, Petal.Width) # by name
select(iris, 5, 4)  # by column index
```


**base R**

```{r}
iris[, c("Species", "Petal.Width")] # by name
iris[, c(5, 4)]  # by column index
```



# Make new variables (columns)

**tidyverse**

```{r}
mutate(iris, 
       Petal.Ratio = Petal.Length/Petal.Width,
       Sepal.Ratio = Sepal.Length/Sepal.Width)
```


**base R**

```{r}
iris$Petal.Ratio <- iris$Petal.Length/iris$Petal.Width
iris$Sepal.Ratio <- iris$Sepal.Length/iris$Sepal.Width
```



# Extract observations (rows)

**tidyverse**

```{r}
filter(iris, Petal.Width > 0.5 & Species == "setosa")
```


**base R**

```{r}
# Using [,]
iris[iris$Petal.Width > 0.5 & iris$Species == "setosa", ]

# Using subset (works very much like dplyr::filter)
subset(iris, Petal.Width > 0.5 & Species == "setosa")
```



# Arrange observations (rows)

**tidyverse**

```{r}
# descending order of species (alphabetic) followed by ascending order of Petal.Width
arrange(iris, desc(Species), Petal.Width) 
```


**base R**

```{r}
# descending order of species (alphabetic) followed by ascending order of Petal.Width
iris[order(rev(iris$Species), iris$Petal.Width) , ]
```



# Summarise observations (rows)

**tidyverse**

```{r}
# Generic way
summarise(iris, 
          Petal.Length.mean = mean(Petal.Length),
          Petal.Length.sd = sd(Petal.Length),
          Sepal.Length.mean = mean(Sepal.Length),
          Sepal.Length.sd = sd(Sepal.Length))

# Shortcut when same functions applied to same variables 
summarise_at(iris, 
             .vars = c("Petal.Length", "Sepal.Length"), 
             .funs = c("mean", "sd"))
```

(see also variants `summarise_all()` and `summarise_if()`)


**base R**

There are many ways to do this with _base R_, here's a couple of possibilities:

```{r}
# Manually create a data.frame
data.frame(Petal.Length.mean = mean(iris$Petal.Length),
           Petal.Length.sd = sd(iris$Petal.Length),
           Sepal.Length.mean = mean(iris$Sepal.Length),
           Sepal.Length.sd = sd(iris$Sepal.Length))

# Use the "aggregate" function
## Column names might have to be changed afterwards
aggregate(formula = c(Sepal.Length, Petal.Length) ~ 1,
          data = iris, 
          FUN = function(x) c(mean = mean(x), sd = sd(x)))
```



# Combine tables

Consider the following tables

```{r, results = TRUE}
band_members
band_instruments
band_instruments2
```

**tidyverse**

```{r}
# Retain rows with matches in both tables
inner_join(band_members, band_instruments, by = "name")  

# Retain all rows:
full_join(band_members, band_instruments, by = "name")  

# Retain all rows from first table:
left_join(band_members, band_instruments, by = "name")  

# Retain all rows from second table:
right_join(band_members, band_instruments, by = "name")  
```

Columns used for merging can have different names between tables:

```{r}
inner_join(band_members, band_instruments2, by = c("name" = "artist"))
```


**base R**

```{r}
# Retain rows with matches in both tables
merge(band_members, band_instruments, by = "name")  

# Retain all rows:
merge(band_members, band_instruments, by = "name", all = TRUE)  

# Retain all rows from first table:
merge(band_members, band_instruments, by = "name", all.x = TRUE)  

# Retain all rows from second table:
merge(band_members, band_instruments, by = "name", all.y = TRUE)  
```

Columns used for merging can have different names between tables:

```{r}
merge(band_members, band_instruments2, by.x = "name", by.y = "artist")
```



# Grouped operations 

**Important note**: with `dplyr`, grouped operations are initiated with the function 
`group_by()`. It is a good habit to use `ungroup()` at the end of a series of 
grouped operations, otherwise the groupings will be carried in downstream analysis, 
which is not always desirable. In the examples below we follow this convention.


**Note on base R**: with base R there are several ways to do these types of operations. 
Here we show a generic "split-apply-combine" solution using the `by()` function. 
These solutions require some understanding of `list` objects and how to write 
custom functions. 


### Summarise rows within groups

**tidyverse**

```{r}
mtcars %>% 
  group_by(cyl, gear) %>% 
  summarise(mpg.mean = mean(mpg),
            mpg.sd = sd(mpg),
            wt.mean = mean(wt),
            wt.sd = sd(wt)) %>% 
  ungroup() # remove any groupings from downstream analysis
```


**base R**

```{r}
# First operate in the data.frame by group (split-apply)
mtcars_by <- by(mtcars, 
   INDICES = list(mtcars$cyl, mtcars$gear),
   FUN = function(x){
     data.frame(cyl = unique(x$cyl),
                gear = unique(x$gear),
                mpg.mean = mean(x$mpg),
                mpg.sd = sd(x$mpg),
                wt.mean = mean(x$wt),
                wt.sd = sd(x$wt))
   })

# Then combine the results into a data.frame
do.call(rbind, mtcars_by)
```

Alternative solution:

```{r}
# Using aggregate
aggregate(formula = cbind(mpg, wt) ~ cyl + gear, 
          data = mtcars, 
          FUN = function(x){
            c(mean = mean(x), sd = sd(x))
          })
```


### Create new columns with calculations done within groups

For example, center the measurements of "Petal.Width" by subtracting the mean 
within species.

**tidyverse**

```{r}
iris %>% 
  group_by(Species) %>% 
  mutate(Petal.Width.centered = Petal.Width - mean(Petal.Width)) %>% 
  ungroup() # remove any groupings from downstream analysis
```


**base R**

```{r}
# First operate in the data.frame by group (split-apply)
iris_by <- by(iris, 
              INDICES = iris$Species, 
              FUN = function(x){
                x$Petal.Width.centered <- x$Petal.Width - mean(x$Petal.Width)
                return(x)
              })

# Then combine the results into a data.frame
do.call(rbind, iris_by)
```

Alternative solution:

```{r}
iris$Petal.Width.centered <- ave(iris$Petal.Width, iris$Species, FUN = function(x) x - mean(x))
```



### Filter rows with conditions evaluated within groups

`iris` flowers with maximum "Petal.Width" for each "Species".

**tidyverse** 

```{r}
iris %>% 
  group_by(Species) %>% 
  filter(Petal.Width == max(Petal.Width))
```

**base R**

```{r}
# First operate in the data.frame by group (split-apply)
widest_petals <- by(iris, 
                    INDICES = iris$Species, 
                    FUN = function(x){
                      x[x$Petal.Width == max(x$Petal.Width), ] 
                    })

# Then combine the results into a data.frame
do.call(rbind, widest_petals)
```




# Reshaping data 

### to "long" format

**tidyverse**

```{r}
gather(iris, key = "trait", value = "measurement", Sepal.Length:Petal.Width)
```


**base R**

I couldn't find an easy way to do this in a clean way, although `reshape()` 
gets us close (but the "trait" variable is coded numerically):

```{r}
reshape(iris, 
        varying = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"),
        timevar = "trait",
        idvar = "id",
        v.names = "measurement",
        direction = "long")
```

There is also `stack()`, but we loose the "Species" column in the process.

```{r}
stack(iris, select = c("Sepal.Length", "Sepal.Width", "Petal.Length", "Petal.Width"))
```



### to "wide" format

**tidyverse**

```{r}
spread(Indometh, key = "time", value = "conc")
```

Note that in this case it might have been better to first modify the "key" column
so that the resulting column names are not numeric:

```{r}
Indometh %>% 
  mutate(time = paste("conc", time, sep = "_")) %>% 
  spread(key = "time", value = "conc")
```


**base R**

```{r}
reshape(Indometh, 
        v.names = "conc",
        idvar = "Subject",
        timevar = "time",
        direction = "wide")
```

