---
title: "Functions and loops - exercise solution"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: pygments
    css: ../assets/styles.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here("data_structures_functions_loops"))

# Function to read and tidy gapminder data
read_gapminder <- function(file){
  
  # Read the data in the file - starts at line 3
  country_data <- read_csv(file, skip = 2, show_col_types = FALSE)
  
  # Extract the world_region from the file - first line of the file
  # and remove the word "Region:" from it
  country_region <- read_lines(file, n_max = 1) %>% 
    str_remove("Region:")

  # Get the country name - from the file name
  country_name <- file %>% basename() %>% str_remove(".csv")
  
  # Add country and world region to our table
  country_data <- country_data %>% 
    mutate(country = country_name, world_region = country_region)
  
  # Return the tidy data
  return(country_data)
}
```

We were given 193 text files, each containing data for a country. 
The format of each text file is the same:

- The name of the file gives the country name
- The first line gives the world region they occur in
- The rest of the data starts at line 3 of the file, and is in a comma-separated format

We were asked to parse these files and combine them into a tidy tabular format. 

There are many ways to solve this exercise. Here, we show a solution using several functions from the `tidyverse` meta-package.

```{r, message = FALSE}
# load the package
library(tidyverse)
```


## Prelude

For any problem where we need to repeat the same task several times, it helps to break down the problem into parts. 
We can write down what the different steps are using words and not worrying about the code.

Here's an example for our problem of parsing and combining the data files:

- For each file make a table with 4 columns: `country`, `world_region`, `year`, `income_per_person`
    - store the file name, which contains the country name
    - extract the first line, which contains the world region
    - extract lines 3 onwards, which contain the year and income
- Repeat the steps above for all files
- Combine all tables together into a single table

Now that we've broken our problem into parts, let's tackle them in turn.

## Understanding our data

If we open one of our files in a text editor, this is how they look like (here showing the first few lines only):

```{r, echo=FALSE}
cat(head(read_lines("data/multiple_files/Afghanistan.csv")), sep = "\n")
```

There are 3 pieces of information that we should bring together into a single 
tidy _data frame_:

* name of the country - this is contained in the file name
* information about the world region - in the first line of the file 
* the rest of the data for the country

That is, we should aim for something like this:

```{r, echo = FALSE}
read_gapminder("data/multiple_files/Afghanistan.csv") %>% 
  select(country, world_region, everything()) %>% 
  head() %>% 
  knitr::kable()
```


## Building the steps to read and tidy our data

We start by focusing on a single file, and then generalise across all files.

If you look at the help of `?read_csv`, you will notice that these `read_*` functions have an option to skip some of the lines of the file. 
This is very convenient, as it allows us to read data starting from line 3, which contain the bulk of the information for our country:

```{r, message=FALSE}
# read the table starting from line 3
read_csv("data/multiple_files/Afghanistan.csv", skip = 2)
```

So this solves one of our steps! 
(note that base R functions such as `read.table()` also have an option to skip lines of a file before reading it in)

Let's now turn into reading the first line of the file, which contains the diet identifier. 
Again, if you look at the help of `read_*` functions, you will notice there is an option `n_max`, which determines how many lines of the file we want to read in (the default is to read all lines). 
We can therefore use this option to read the first line of the file only.

We will use a function called `read_lines()`, which simply reads the lines of a file and stores them into a vector. 
This function is not often used, but can be useful when parsing data.

```{r}
# Read the first line of the file
read_lines("data/multiple_files/Afghanistan.csv", n_max = 1)
```

Because we are reading the first line of the file only, the output is a vector of length 1. 
For our final dataset, we don't really need to keep the string "Region:" because we will place these values in a column. 
Therefore, it is useful to remove that string from our value. 
We will use the function `str_remove()`, which we can pipe from the previous line of code:

```{r}
# Read first line and remove "Diet:" string
read_lines("data/multiple_files/Afghanistan.csv", n_max = 1) %>% 
  str_remove("Region:")
```

And that's another problem solved!

Finally, we want to extract the country name from the file name. 
So far we've been looking at a single file, but eventually we will create a list of file names too  apply these steps to. 
For now, let's manually type the file name of the first file we've been experimenting on:

```{r}
"data/multiple_files/Afghanistan.csv"
```

From this file name, we want to extract the file name only (without the directory), 
which we can do using the `basename()` function:

```{r}
"data/multiple_files/Afghanistan.csv" %>% basename()
```

Then, we want to remove the string ".txt" from it, which we've already seen can be done with `str_remove()`:

```{r}
"data/multiple_files/Afghanistan.csv" %>% 
  basename() %>% 
  str_remove(".csv")
```

And that's it!


## Putting it all together into a function

We can wrap all these steps into our own custom function. 
This will be convenient when we want to apply this function to the entire list of files.

```{r}
read_gapminder <- function(file){
  
  # Read the data in the file - starts at line 3
  country_data <- read_csv(file, skip = 2, show_col_types = FALSE)
  
  # Extract the world_region from the file - first line of the file
  # and remove the word "Region:" from it
  country_region <- read_lines(file, n_max = 1) %>% 
    str_remove("Region:")

  # Get the country name - from the file name
  country_name <- file %>% basename() %>% str_remove(".csv")
  
  # Add country and world region to our table
  country_data <- country_data %>% 
    mutate(country = country_name, world_region = country_region)
  
  # Return the tidy data
  return(country_data)
}
```

Let's test the function:

```{r}
read_gapminder("data/multiple_files/Afghanistan.csv")
```



## Apply function to list of files

First let's create a list of data files in our directory:

```{r}
file_list <- list.files("data/multiple_files/", pattern = ".csv",
                        full.names = TRUE)
head(file_list)
```

We can now for example use a for loop to read all the files into a list:

```{r}
# initialise an empty list with the correct length
gapminder <- vector("list", length = length(file_list))

# loop through the files and add to the list
for(i in seq_along(file_list)){
  gapminder[[i]] <- read_gapminder(file_list[i])
}
```

The result is now a _list_ containing `r length(gapminder)` data frames, one of each country. 

Finally, we can "bind" those tables using the `bind_rows()` function:

```{r}
gapminder <- bind_rows(gapminder)
```


:::note 

An alternative to using _for loops_ in R is to use functions that are designed to iterate through vectors/lists.
Base R has the function `lapply()` and tidyverse as the similar function called `map()`. 

These take as an input a vector and a function, and will apply the function to each element of the vector. 

Here is the same task as we did above with the _for loop_ using the `lapply()` function instead: 

```{r, eval=FALSE}
gapminder <- lapply(file_list, read_gapminder)
```

You can think of `map()`/`lapply()` as encapsulated _for loops_. 
The advantage of using these functions is that the code is more compact and there are less intermediate variables cluttering our environment (in this case, the variable `i` was created by the for loop, but no such variable is created by the `map()` function). 

Learn more about `map()` and its variants from the 
[iteration chapter in R for Data Science](http://r4ds.had.co.nz/iteration.html)

:::

