---
title: "Introduction to ggplot2"
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 3
    highlight: pygments
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lesson objectives

This lesson complements the [Data Carpentry lessons]() 
by providing an earlier introduction to `ggplot2`.
The lesson is to be taught after [Starting With Data]() 
but before [Data Manipulation]() lesson. 

* Identify the basic parts needed to build a graph with `ggplot2`.
    * defining data, aesthetics and geometries for a basic graph.
* Distinguish when to use or not to use `aes()` to change graph's aesthetics (e.g. colours, shapes).
* Apply overlaying of multiple geometries on the same graph and define aesthetics separetely for each.
* Adjust and costumise scales and labels in the graph.


```{r, echo=FALSE, message=FALSE, fig.width=7, fig.height=9}
library(patchwork); library(tidyverse)
surveys <- read_csv("https://ndownloader.figshare.com/files/2292169")

surveys_nomiss <- drop_na(surveys)

p1 <- surveys_nomiss %>% 
  ggplot(aes(x = weight, colour = genus)) +
  geom_freqpoly(binwidth = 1) +
  labs(x = "Hindfoot length", title = "Frequency polygon") +
  theme(legend.position = "none")

p2 <- surveys_nomiss %>% 
  ggplot(aes(x = genus, y = weight)) +
  geom_violin(scale = "width", fill = "grey") +
  geom_boxplot(width = 0.2) +
  labs(title = "Boxplot & Violin plot") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1, size = 8))

p3 <- surveys_nomiss %>% 
  ggplot(aes(x = weight, y = hindfoot_length, colour = genus)) +
  geom_point() +
  labs(title = "Scatterplot") +
  theme(legend.position = "bottom")

{p1 + p2 + plot_layout(ncol = 2)} / p3 + plot_layout(heights = c(1, 2))
```



# Setup

Make sure you loaded the `surveys` data:

```{r, eval=FALSE}
library(tidyverse)
surveys <- read_csv("data/portal_data_joined.csv")
```

For this lesson we're going to remove missing values from our table.
This is not always desirable (you might be throwing away good data!), but it will 
help us make this lesson clearer.

We will use the function `drop_na()`, which removes rows with any missing data:

```{r}
# remove rows with missing data in any of the columns
surveys_nomiss <- drop_na(surveys)
```



# Building a `ggplot2` graph

To build a `ggplot2` graph you need 3 basic pieces of information:

* A **data.frame** with data to be plotted
* The variables (columns of `data.frame`) that will be _mapped_ to different 
**aesthetics** of the graph (e.g. axis, colours, shapes, etc.)
* the **geometry** that will be draw on the graph (e.g. points, lines, 
boxplots, violinplots, etc.)

This translates into the following basic syntax:

``` r
ggplot(data = <data.frame>, 
       mapping = aes(x = <column of data.frame>, y = <column of data.frame>)) +
   geom_*()
```


For our first visualisation, let's make a scatterplot showing the relationship 
between `weight` and `hindfoot_length`. 
Let's do it step-by-step to see how `ggplot2` works. 

Start by giving data to `ggplot`:

```{r}
ggplot(data = surveys_nomiss)
```

OK, that "worked" (as in, we didn't get an error). 
But because we didn't give `ggplot()` any variables to be _mapped_ to **aesthetic** 
components of the graph, we just got a grey square. 

For _mappping_ columns to _aesthetics_, we use the `aes()` function to give this 
information for the axis:

```{r}
ggplot(data = surveys_nomiss, 
       mapping = aes(x = weight, y = hindfoot_length))
```

That's better, now we have some axis. Notice how `ggplot()` defines the axis based on the range of data given.
But it's still not a very interesting graph, because we didn't tell what it is we want to draw on the graph.

This is done by adding (literally `+`) _geometries_ to our graph:

```{r}
ggplot(data = surveys_nomiss, 
       mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point()
```

----

**Exercise**

> * Modify the graph above by plotting a density hexagon plot (`geom_hex()`)
> * From this graph, we can see that there are different groups of observations.
> Either alone or with person next to you discuss why this might be and how you would 
> like to change the graph to investigate it. 

----


### Changing how geometries look like

We can change how geometries look like in several ways, for example their transparency, colour, shape, etc. 

To know which _aesthetic_ components can be changed in a particular geometry, 
look at its help (e.g. `?geom_point`) and look under the "Aesthetics" section of the help page. 

For example, because the points in the above graph are quite densely packed, 
we can change the transparency of the points in our scatterplot using `alpha`
(`alpha` varies between 0-1 with zero being transparent and 1 being opaque):

```{r}
ggplot(data = surveys_nomiss, 
       mapping = aes(x = weight, y = hindfoot_length)) +
  geom_point(alpha = 0.1)
```

With this transparency we can see which areas of the graph are more densely occupied with points. 

----

**Exercise**

> Try changing the _size_ and _shape_ of the lines 
> (hint: web search "point shape in R" to see how to make a triangle)

----


### Changing aesthetics based on data

In the above exercise we changed the colour of the points by defining it ourselves. 
However, it would be better if we coloured the points based on a variable of interest. 

For example, it's likely that the clustering of points in our scatterplot is due to differences between genera. 

We do this by passing this information to the `colour` _aesthetic_ (inside the `aes()` function):

```{r}
ggplot(data = surveys_nomiss, 
       mapping = aes(x = weight, y = hindfoot_length, colour = genus)) +
  geom_point()
```

This illustrates an important distinction between _aesthetics_ defined inside or outside of `aes()`:

* if you want the _aesthetic_ to change **based on the data** it goes **inside `aes()`**
* if you want to **manually specify** how the geometry should look like, it goes **outside `aes()`**

----

**Exercise**

> Make a boxplot showing the distribution of _weight_ (y-axis) for each `genus` (x-axis).
> (hint: use the `geom_boxplot()` function)
> Bonus: Colour the inside of the boxplots by sex 

----

### Multiple geometries

Often, we may want to overlay several geometries on top of each other. 
For example, we could use a violin plot as an alternative to a boxplot: 

```{r}
ggplot(surveys_nomiss, aes(x = genus, y = weight)) +
  geom_violin(scale = "width")
```

But even better, we may want to overlay a boxplot on top of each violinplot, 
so that we get both representations of the data. 
We do this by "adding" (with `+`) another geometry to the graph:

```{r}
ggplot(surveys_nomiss, aes(x = genus, y = weight)) +
  geom_violin(scale = "width") +
  geom_boxplot(width = 0.2)
```

The order in which you add the geometries defines the order they are "drawn" on the graph. 
For example, try swapping their order and see what happens 
(bonus: try adding transparency to the violin plot)


### Controlling aesthetics in individual geometries

Let's say that, in the graph above, we wanted to colour the violins by genus, but keep the boxplots without colour. 

Because we want to colour our geometries _based on data_, this goes inside the `aes()` part of the graph:

```{r}
ggplot(surveys_nomiss, aes(x = genus, y = weight, fill = genus)) +
  geom_violin(scale = "width") +
  geom_boxplot(width = 0.2)
```

OK, this is not what we wanted. Both _geometries_ (boxplots and violins) got coloured. 

It turns out that we can control aesthetics individually in each geometry, using the `aes()` function.
Like this:

```{r}
ggplot(surveys_nomiss, aes(x = genus, y = weight)) +
  geom_violin(scale = "width", aes(fill = genus)) +
  geom_boxplot(width = 0.2)
```

----

**Exercise**

> Modify the graph above by colouring the inside of the boxplots by genus and the inside of the violins in grey colour. 

```{r, echo=FALSE}
ggplot(surveys_nomiss, aes(x = genus, y = weight)) +
  geom_violin(scale = "width", fill = "grey") +
  geom_boxplot(width = 0.2, aes(fill = genus))
```

----


### Modifying scales

Often you want to change how the scales of your plot are defined. 
In `ggplot2` scales can refer to the `x` and `y` aesthetics, but also to other 
aesthetics such as `colour`, `shape`, `fill`, etc. 

We modify scales using the _scale_ family of functions. 
These functions always follow the following naming convention: `scale_<aesthetic>_<type>`, where:

* `<aesthetic>` refers to the aesthetic for that scale function (e.g. `x`, `y`, `colour`, `fill`, `shape`, etc.)
* `<type>` refers to the type of aesthetic (e.g. `discrete`, `continuous`, `manual`)

Let's see some examples.

Taking the graph from the previous exercise we modify the y-axis scale to emphasise 
the lower weights of our animals:

```{r}
ggplot(surveys_nomiss, aes(x = genus, y = weight)) +
  geom_violin(scale = "width", fill = "grey") +
  geom_boxplot(width = 0.2, aes(fill = genus)) +
  scale_y_continuous(limits = c(0, 100))
```

Another example: change the `fill` scale to define custom colours "manually". 
This is a "discrete" type of scale (our values are categorical, not continuous):

```{r}
ggplot(surveys_nomiss, aes(x = genus, y = weight)) +
  geom_violin(scale = "width", fill = "grey") +
  geom_boxplot(width = 0.2, aes(fill = genus)) +
  scale_fill_manual(values = c("pink", "orange", "green4", "blue4", "purple", "tomato", "brown", "grey", "black"))
```

For colour/fill scales there's a very convenient variant of the function ("brewer") that has some pre-defined palettes:

```{r}
ggplot(surveys_nomiss, aes(x = genus, y = weight)) +
  geom_violin(scale = "width", fill = "grey") +
  geom_boxplot(width = 0.2, aes(fill = genus)) +
  scale_fill_brewer(palette = "Set1")
```

Look at the function's help documentation for more information on palettes. 
Note that some palettes only have a limited number of colours and `ggplot` will 
give a warning. 


#### In summary

* To build a `ggplot2` graph you need to provide _data_, _aesthetics_ and _geometries_.
* If you want to change an _aesthetic_ of our graph **based on data**, include it **inside `aes()`**.
* If you want to **manually** change an _aesthetic_ regardless of data then it goes **outside `aes()`**.
* You can overlay multiple geometries in the same graph, and control their aesthetics individually.
* You can adjust scales of your graph using `scale_*` family of functions.


# List of graphs for exploratory data analysis

Here we just quickly showcase different types of graphs that are suitable for 
different types of questions you may want to ask from your data. 

### Variation of a continuous variable

[density plot](https://www.data-to-viz.com/graph/density.html) or
[histograms](https://www.data-to-viz.com/graph/histogram.html) 


```{r}
ggplot(data = surveys_nomiss, 
       mapping = aes(x = weight)) +
  geom_density(fill = "grey")
```


### Variation of a continuous variable across categories

[boxplot](https://www.data-to-viz.com/caveat/boxplot.html) or 
[violin](https://www.data-to-viz.com/graph/violin.html)

```{r}
ggplot(surveys_nomiss, 
       aes(x = genus, y = weight)) +
  geom_boxplot()
```

[Ridge plots](https://www.data-to-viz.com/graph/ridgeline.html) can be particularly useful if 
your y-axis is ordered (which is not the case here, but still useful)

```{r}
library(ggridges)
ggplot(surveys_nomiss, 
       aes(x = weight, y = genus)) +
  geom_density_ridges()
```


### Relationship between two continuous variables 

[scatterplot](https://www.data-to-viz.com/graph/scatter.html)

```{r}
ggplot(data = surveys_nomiss, 
       mapping = aes(x = weight, y = hindfoot_length, colour = genus)) +
  geom_point()
```

[2D density plot](https://www.data-to-viz.com/graph/density2d.html)

```{r}
ggplot(data = surveys_nomiss, 
       mapping = aes(x = weight, y = hindfoot_length)) +
  geom_hex()
```

