---
title: "Recap Day 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Create objects

In R **create objects** (something that contains a value) with `<-`

```{r}
x <- 53.341
```

### Functions

We can use **functions** to do specific operations

```{r}
round(x)   # round the the value of x
```

Functions have **options** that can change their behaviour.
Separate options using a comma

```{r}
round(x, digits = 1) # round to one decimal point
```


### Vector

A **vector** contains several values _all of the same type_
(numeric, character, logical)

```{r}
x_chr <- c("dog", "cat", "goldfish")   # character vector
x_num <- c(1, 5, 23.3, 55.2)           # numeric vector
x_log <- c(TRUE, TRUE, FALSE, TRUE)    # logical vector
```

**Access values** inside the vector with `[]`

```{r}
x_chr[2]
x_num[c(2, 4)]
```


### Conditions

Create logical vectors using **conditions**

```{r}
x_num
x_num > 20                # is x_num greater than 20
x_num == 5                # is x_num equal to 5
x_num %in% c(20, 30, 1)   # is x_num contained the vector on the right
```

Combine conditions with & (AND) and | (OR)

```{r}
x
x_num >= 10 & x_num <= 30   # is x_num greater than or equal to 10 AND smaller than or equal to 30
x_num < 10 | x_num > 30   # is x_num smaller than 10 OR greater than 30
```

We can use logical vectors (TRUE/FALSE) to subset vectors

```{r}
x_num[x_num > 20]   # return values of x_num where x_num is greater than 20
```

### data.frame

A **data.frame** is a tabular object (rows and columns).
Usually we create these from a file with `read.csv()`

```{r}
# Add option to turn off behaviour of converting strings to factors
surveys <- read.csv("https://ndownloader.figshare.com/files/2292169",
                    stringsAsFactors = FALSE)
```

Subset data.frames using **[rows, columns]**

```{r}
surveys[1:6, c(5, 11)]   # rows 1 to 6 and columns 5 and 11
```

Access individual columns by name using **`$`**
```{r, eval = FALSE}
surveys$species   # returns a vector with the values of the species column
```

```{r, echo = FALSE}
c(surveys$species[1:10], "...")
```

### missing values

Sometimes we have **missing values**, encoded as `NA`

```{r}
y <- c(23, 44, NA, 212)
```

We need to ensure these are dealt with properly

```{r}
mean(y)   # returns NA
mean(y, na.rm = TRUE)  # removes NA and then calculates mean
```

The `is.na()` function is important to deal with missing values:

```{r}
y

# create a logical that is true if value is missing
is.na(y)

# Negate that expression using !
!is.na(y)
```

We can **remove NA** by using this function:

```{r}
y[!is.na(y)]  # return values of y that are not missing
```




























